Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-08 pldi-2016/paper.tex
More edits in section 4 and 5

\begin{figure*}
\noindent
\begin{equation*}
\hspace{1.1em}
\begin{array}{l}
\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
\\[0.5em]
\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
\end{array}  
\quad
\begin{array}{l}
\\[-2.5em]
\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
\end{array}
\end{equation*}
%
\vspace{-0.75em}
%
\begin{equation*}
\begin{array}{l}
\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
\end{array}
\;\;
\begin{array}{l}
\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
\\[0.25em]
\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
\\[0.25em]
\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
\end{array}
\end{equation*}

\caption{Reduction rules for conversion functions}
\label{fig:op-conversions}
\vspace{-0.5em}
\end{figure*}

@@

The behaviour of the dynamic data operations is defined by the reduction rules in 
Figure~\ref{fig:op-conversions}. The typing is shown in Figure~\ref{fig:ff-typecheck} and is 
discussed later. The \ident{hasShape} function represents a runtime shape test.
It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
argument has a shape specified by the first argument. For records, we have to check that for each 
field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name 
with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false} 
for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for 
\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo 
expression.

@@

@@

Figure~\ref{fig:ff-reduction} shows the reduction rules.  The (\emph{member}) rule reduces a member access using a class 
definition in the assumption. The (\emph{ctx}) rule models the eager
evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
context $E$:
%
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
\end{array} 
\end{equation*}

@@

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.6em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
 { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq1)}}&
 v=v'\reduce\kvd{true} \qquad (\textnormal{when}~v = v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq2)}}&
 v=v'\reduce\kvd{false} \qquad (\textnormal{when}~v \neq v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(fun)}}&
 \hspace{-0.4em}
 (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
 \\[-0.3em]
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

\caption{Foo -- Remaining reduction rules}
\label{fig:ff-reduction}
\end{figure}

@@

\begin{figure}
\noindent  
\begin{equation*}
\begin{array}{c}  
\inference
  {~}
  {L; \Gamma \vdash d : \ident{Data}}
~~
\inference
  {~}
  {L; \Gamma \vdash i : \ident{int}}
~~
\inference
  {~}
  {L; \Gamma \vdash f : \ident{float}}
\\[1.75em]
\inference
  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
  {L; \Gamma \vdash \lambda x.e : \tau_2}
\quad
\inference
  {L; \Gamma \vdash e_2 : \tau_1 \\
   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
  {L; \Gamma \vdash e_1~e_2 : \tau_2}
\\[1.75em]
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\\[1.75em]
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\\[1.75em]
\inference
  {L; \Gamma \vdash e : \ident{Data}}
  {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
\\[1.75em]
\inference
  { L; \Gamma \vdash e : \tau & \tau \in \{ \ident{int}, \ident{float} \} }
  { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
\\[1.75em]
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convNull}(e_1, e_2) : \ident{option}\langle\tau\rangle }
\\[1.75em]
\inference
  { L; \Gamma \vdash e : \ident{Data} \\
    (\sigma,\tau)\in\{ (\ident{int},\ident{int}), (\ident{string},\ident{string}), (\ident{bool},\ident{bool}) \} }
  { L; \Gamma \vdash \ident{convPrim}(\sigma, e) : \tau }
\\[1.75em]
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
\\[1.75em]
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
\end{array}
\end{equation*}

\caption{Foo -- Fragment of type checking}
\label{fig:ff-typecheck}
\end{figure}

@@

\paragraph{Example 1.}
To illustrate how the mechanism works, we consider two examples. First, assume 
that the inferred shape is a record  
$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
The rules from Figure~\ref{fig:tp-generation} produce the following class (we reduce some 
function applications for readability):
%
\vspace{-0.1em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(x_1:\ident{Data})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \ident{convField}(\ident{Person}, \ident{Age}, x_1, \lambda x_2 \rightarrow \\[0.1em]
 \qquad \quad \ident{convNull}(x_2, \lambda x_3\rightarrow\ident{convPrim}(\ident{int}, x_3))~) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{convField}(\ident{Person},\ident{Name}, x_1, \lambda x_2 \rightarrow \\[0.1em]
 \qquad \quad \ident{convPrim}(\ident{string}, x_2)))
\end{array}
\end{equation*}
%
The body of the \ident{Age} member uses \ident{convField} as specified by the case for optional
record fields. The field shape is nullable and so \ident{convNull} is used in the continuation to 
convert the value to \ident{None} if \ident{convField} produces a \kvd{null} data value and 
\ident{hasShape} is used to ensure that the field has the correct shape. The \ident{Name} value should
be always available and should have the right shape so \ident{convPrim} appears directly in the 
continuation. This is where the evaluation can get stuck if the field value was missing. The F\# 
type corresponding to the record shape is \ident{Person} the function to create it from a data value 
is $\lambda x . \kvd{new}~\ident{Person}(x)$.

\paragraph{Example 2.} The second example illustrates the handling of collections and 
labelled top types. Reusing \ident{Person} from the previous example, consider 
$[\kvd{any}\langl\ident{Person}~\{ \ldots \},\ident{string}\rangl]$:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{PersonOrString}(x:\ident{Data})~= \\[0.1em]
 \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{hasShape}(\ident{Person}~\{ \ldots \}, x)~\kvd{then}\\[0.1em]
 \qquad\quad \ident{Some}(\kvd{new}~\ident{Person}(x))~\kvd{else}~\ident{None} \\[0.1em]
 \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{hasShape}(\ident{string}, x)~\kvd{then}\\[0.1em]
 \qquad\quad \ident{Some}(\ident{convPrim}(\ident{string}, x))~\kvd{else}~\ident{None}
\end{array}
\end{equation*}
%
The type provider maps the collection of labelled top shapes to a type $\ident{list}\langl\ident{PersonOrString}\rangl$
and returns a function that parses a data value as follows:
%
\begin{equation*}
\lambda x_1\rightarrow \ident{convElements}(x_1\lambda x_2\rightarrow\kvd{new}~\ident{PersonOrString}(x_2))
\end{equation*}
%
The \ident{PersonOrString} class contains one member for each of the labels. In the body, they
check that the input data value has the correct shape using \ident{hasShape}. This also implicitly 
handles \kvd{null} by returning \kvd{false}. As discussed earlier, labelled top types provide easy 
access to the known cases (\ident{string} or \ident{Person}), but they require a runtime shape check.

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
\item[--] Records in the input can have additional fields
\item[--] Records in the input can have fewer fields, provided that the type of the field is nullable in some of the samples
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption  
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
provided expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

@@

\noindent
This shows that provided types are correct with respect to the preferred shape relation. 
This is an obligation upon the author of a type provider, which is a second reason why
the type soundness is ``relative''.
Our key theorem states that, for any input which is a subshape the inferred shape and 
any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
(reduction does not change type) and progress (an expression can be reduced).

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

\begin{lemma}[Progress]
\label{thm:rs-progress}
Given the assumptions and definitions from Theorem~\ref{thm:safety}, there exists $e''$ such that
$e[y\leftarrow e'~d'] \reduce e''$.
\end{lemma}
\begin{proof}
Proceed by induction over the typing derivation of $L; \emptyset \vdash e[y\leftarrow e'~d'] : \tau'$. 
The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.
\end{proof}

>>>>

\begin{figure*}
\noindent
\begin{center}
\textbf{Part I.} Reduction rules for conversion functions
\end{center}

\vspace{0.5em}
\noindent
\begin{equation*}
\begin{array}{l}
\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
\\[0.5em]
\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
\end{array}  
\quad
\begin{array}{l}
\\[-2.5em]
\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
\end{array}
\end{equation*}
%
\vspace{-0.75em}
%
\begin{equation*}
\hspace{-1.1em}
\begin{array}{l}
\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
\end{array}
\;\;
\begin{array}{l}
\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
\\[0.25em]
\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
\\[0.25em]
\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
\end{array}
\end{equation*}

\noindent
\begin{center}
\textbf{Part II.} Reduction rules for the rest of the Foo calculus 
\end{center}

\vspace{0.5em}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.6em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
 { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq1)}}&
 v=v'\reduce\kvd{true} \qquad (\textnormal{when}~v = v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq2)}}&
 v=v'\reduce\kvd{false} \qquad (\textnormal{when}~v \neq v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(fun)}}&
 \hspace{-0.4em}
 (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
\end{array}
\quad
\begin{array}{rl}
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
\\[-0.3em]
\textnormal{\footnotesize{(ctx)}}&
\hspace{-0.4em}
E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

\caption{Foo -- Reduction rules for the Foo calculus and dynamic data operations}
\label{fig:ff-reduction}
\end{figure*}

@@

The behaviour of the dynamic data operations is defined by the reduction rules in 
Figure~\ref{fig:ff-reduction} (Part I). The typing is shown in Figure~\ref{fig:ff-typecheck} and is 
discussed later. The \ident{hasShape} function represents a runtime shape test.
It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
argument has a shape specified by the first argument. For records, we have to check that for each 
field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name 
with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false} 
for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for 
\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo 
expression.

@@

\begin{figure*}
\noindent  
\begin{equation*}
\begin{array}{c}  
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash d : \ident{Data}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash i : \ident{int}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash f : \ident{float}\hspace{-0.5em}}
~~
\inference
  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
  {L; \Gamma \vdash \lambda x.e : \tau_2}
~~
\inference
  {L; \Gamma \vdash e_2 : \tau_1 &
   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
  {L; \Gamma \vdash e_1~e_2 : \tau_2}
\\[1.75em]
\inference
  {L; \Gamma \vdash e : \ident{Data}}
  {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
\quad
\inference
  { L; \Gamma \vdash e : \tau & \tau \in \{ \ident{int}, \ident{float} \} }
  { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convNull}(e_1, e_2) : \ident{option}\langle\tau\rangle }
\\[1.75em]
\inference
  { L; \Gamma \vdash e : \ident{Data} \\
    \ident{prim}\in\{ \ident{int}, \ident{string}, \ident{bool} \} }
  { L; \Gamma \vdash \ident{convPrim}(\ident{prim}, e) : \ident{prim} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
\\[1.75em]
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\qquad
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{array}
\end{equation*}

\caption{Foo -- Fragment of type checking}
\label{fig:ff-typecheck}
\end{figure*}

@@

Figure~\ref{fig:ff-reduction} (Part II) shows the reduction rules.  The (\emph{member}) rule reduces 
a member access using a class definition in the assumption. The (\emph{ctx}) rule models the eager
evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
context $E$:
%
\begin{equation*}
\begin{array}{rcl}
 E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
\end{array} 
\end{equation*}

@@

@@

@@

\paragraph{Example 1.}
To illustrate how the mechanism works, we consider two examples. First, assume 
that the inferred shape is a record  
$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
The rules from Figure~\ref{fig:tp-generation} produce the following class \ident{Person} (shown below)
with two members.

The body of the \ident{Age} member uses \ident{convField} as specified by the case for optional
record fields. The field shape is nullable and so \ident{convNull} is used in the continuation to 
convert the value to \ident{None} if \ident{convField} produces a \kvd{null} data value and 
\ident{hasShape} is used to ensure that the field has the correct shape. The \ident{Name} value should
be always available and should have the right shape so \ident{convPrim} appears directly in the 
continuation. This is where the evaluation can get stuck if the field value was missing:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Person}(x_1:\ident{Data})~= \\[0.1em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
 \qquad \ident{convField}(\ident{Person}, \ident{Age}, x_1, \lambda x_2 \rightarrow \\[0.1em]
 \qquad \quad \ident{convNull}(x_2, \lambda x_3\rightarrow\ident{convPrim}(\ident{int}, x_3))~) \\[0.1em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
 \qquad \ident{convField}(\ident{Person},\ident{Name}, x_1, \lambda x_2 \rightarrow \\[0.1em]
 \qquad \quad \ident{convPrim}(\ident{string}, x_2)))
\end{array}
\end{equation*}
%
The function to create the Foo value \ident{Person} from a data value 
is $\lambda x . \kvd{new}~\ident{Person}(x)$.

\paragraph{Example 2.} The second example illustrates the handling of collections and 
labelled top types. Reusing \ident{Person} from the previous example, consider 
$[\kvd{any}\langl\ident{Person}~\{ \ldots \},\ident{string}\rangl]$:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{PersonOrString}(x:\ident{Data})~= \\[0.1em]
 \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{hasShape}(\ident{Person}~\{ \ldots \}, x)~\kvd{then}\\[0.1em]
 \qquad\quad \ident{Some}(\kvd{new}~\ident{Person}(x))~\kvd{else}~\ident{None} \\[0.1em]
 \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
 \qquad \kvd{if}~\ident{hasShape}(\ident{string}, x)~\kvd{then}\\[0.1em]
 \qquad\quad \ident{Some}(\ident{convPrim}(\ident{string}, x))~\kvd{else}~\ident{None}
\end{array}
\end{equation*}
%
The type provider maps the collection of labelled top shapes to a type $\ident{list}\langl\ident{PersonOrString}\rangl$
and returns a function that parses a data value as follows:

\noindent
\begin{equation*}
\lambda x_1\rightarrow \ident{convElements}(x_1\lambda x_2\rightarrow\kvd{new}~\ident{PersonOrString}(x_2))
\end{equation*}
%
The \ident{PersonOrString} class contains one member for each of the labels. In the body, they
check that the input data value has the correct shape using \ident{hasShape}. This also implicitly 
handles \kvd{null} by returning \kvd{false}. As discussed earlier, labelled top types provide easy 
access to the known cases (\ident{string} or \ident{Person}), but they require a runtime shape check.

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
\item[--] Records in the input can have additional fields
\item[--] Records in the input can have fewer fields than some of the records in the sample 
  document, provided that the sample also contains records that do not have the field 
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption  
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
provided expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

@@

\noindent
This shows that provided types are correct with respect to the preferred shape relation. 
Our key theorem states that, for any input which is a subshape the inferred shape and 
any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
(reduction does not change type) and progress (an expression can be reduced).

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

\begin{lemma}[Progress]
\label{thm:rs-progress}
Given the assumptions and definitions from Theorem~\ref{thm:safety}, there exists $e''$ such that
$e'[y\leftarrow e~d'] \reduce e''$.
\end{lemma}
\begin{proof}
Proceed by induction over the typing derivation of $L; \emptyset \vdash e[y\leftarrow e'~d'] : \tau'$. 
The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.
\end{proof}
